Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-10-13 article.tex
Made section 5.4's relationshipto table 4 more obvious
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1401 ccdd05d5-56c1-461a-bf98-85f6876523b4

The conventional, direct \TD\ method consists of a bank of \fir\ filters, or
sliding-window dot products.  If there are $\numtmps$ templates, each
$\tmpsamps$ samples in length, then each filter requires $M N$ multiplications
and additions per sample, or $2 \numtmps \tmpsamps f^0$ \flops\ at a sample rate
$f^0$.

@@

Modern implementations of the \fft, such as the ubiquitous \texttt{fftw}, require about
$2 \fftblock \lg \fftblock$ operations to evaluate a real transform of size
$\fftblock$~\citep{Johnson:2007p9654}.  Including the forward transform of the data and
$M$ reverse transforms for each of the templates, the \fft\ costs $2 (\numtmps + 1)
\fftblock \lg \fftblock$ operations per block.  The multiplication of the transforms adds
a further $2 \numtmps \fftblock$ operations per block.  Since each block produces
$\fftblock - \tmpsamps$ usable samples of output, the overlap-save method requires
$$
f^0 \cdot \frac{2 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock} \; \mathrm{\flops} \,.
$$

@@

Table~\ref{table:flops} shows that the \lloid\ method requires $6.6 \times 10^8$
\flops\ to cover a sub-bank comprising 657 out of the total 98544 mass pairs.
Assuming that other regions of the parameter space have similar computational
scaling, an entire single detector search for \NS--\NS\ signals in the
1--3~$M_\odot$ component mass range could be implemented with
$\sim$$10^{11}$~\flops.  Modern (ca. 2011) workstations can achieve peak
computation rates up to $\sim$$10^{11}$~\flops.  In practice, we expect that a
software implementation of \lloid\ will reach average computation rates that are
perhaps a factor 10 less than this, $\sim$$10^{10}$~\flops\ per machine, due to
non-floating point tasks including bookkeeping and thread synchronization.
Given these considerations, we estimate that a full single-detector Advanced
\LIGO\ search with \lloid\ will require about $\sim$$10$ machines.

We computed the computational cost of a full Advanced \LIGO\ \NS--\NS\ search a
second way by dividing the entire 1--3~$M_\odot$ parameter space into sub-banks
of 657 points apiece, performing time slices and SVDs for each sub-bank, and
tabulating the number of floating point operations using
expression~(\ref{eq:lloid-flops}).  This should be a much more accurate measure
because template length varies over the parameter space.  Lower chirp mass
templates sweep through frequency more slowly and require more computations
while higher chirp mass templates are shorter and require fewer computations.
Despite these subtleties, this estimate gave us $1.1 \times 10^{11}$~\flops,
agreeing with the simple scaling argument above.  Therefore we believe that
$\sim$$10$ machines is a realistic estimate of our computing requirements.

By comparison, using the conventional \TD\ method to achieve the same latency costs
$\sim$$10^{13}$ \flops\ for this particular sub-bank, and so it would require
a total of $\gtrsim$$10^{15}$ \flops\ or $\gtrsim$$10^5$ present-day machines
to search the full parameter space.  Presently, the \LIGO\ Data Grid%
\footnote{\url{https://www.lsc-group.phys.uwm.edu/lscdatagrid/}} consists of
only $\sim$$10^4$ machines, so direct \TD\ convolution is clearly impractical.

>>>>

The conventional, direct \TD\ method consists of a bank of \fir\ filters, or
sliding-window dot products.  If there are $\numtmps$ templates, each
$\tmpsamps$ samples in length, then each filter requires $M N$ multiplications
and additions per sample, or, at a sample rate $f^0$,
%
\begin{equation}
	\label{eq:td-flops}
	2 \numtmps \tmpsamps f^0 \textrm{ \flops}.
\end{equation}

@@

Modern implementations of the \fft, such as the ubiquitous \texttt{fftw}, require about
$2 \fftblock \lg \fftblock$ operations to evaluate a real transform of size
$\fftblock$~\citep{Johnson:2007p9654}.  Including the forward transform of the data and
$M$ reverse transforms for each of the templates, the \fft\ costs $2 (\numtmps + 1)
\fftblock \lg \fftblock$ operations per block.  The multiplication of the transforms adds
a further $2 \numtmps \fftblock$ operations per block.  Since each block produces
$\fftblock - \tmpsamps$ usable samples of output, the overlap-save method requires
%
\begin{equation}
	\label{eq:fd-flops}
	f^0 \cdot \frac{2 (\numtmps + 1) \lg \fftblock + 2 \numtmps}{1 - \tmpsamps/\fftblock} \textrm{ \flops}.
\end{equation}

@@

Table~\ref{table:flops} shows that the \lloid\ method requires $6.6 \times 10^8$
\flops\ to cover a sub-bank comprising 657 out of the total 98544 mass pairs.
Assuming that other regions of the parameter space have similar computational
scaling, an entire single detector search for \NS--\NS\ signals in the
1--3~$M_\odot$ component mass range could be implemented with $(98544/657)\approx150$ times the cost, or $9.9 \times 10^{10}$~\flops.

We computed the computational cost of a full Advanced \LIGO\ \NS--\NS\ search a
second way by dividing the entire 1--3~$M_\odot$ parameter space into sub-banks
of 657 points apiece, performing time slices and SVDs for each sub-bank, and
tabulating the number of floating point operations using
expression~(\ref{eq:lloid-flops}).  This should be a much more accurate measure
because template length varies over the parameter space.  Lower chirp mass
templates sweep through frequency more slowly and require more computations
while higher chirp mass templates are shorter and require fewer computations.
Despite these subtleties, this estimate gave us $1.1 \times 10^{11}$~\flops,
agreeing with the simple scaling argument above.

Modern (ca. 2011) workstations can achieve peak
computation rates up to $\sim$$10^{11}$~\flops.  In practice, we expect that a
software implementation of \lloid\ will reach average computation rates that are
perhaps a factor 10 less than this, $\sim$$10^{10}$~\flops\ per machine, due to
non-floating point tasks including bookkeeping and thread synchronization.
Given these considerations, we estimate that a full single-detector Advanced
\LIGO\ search with \lloid\ will require $\sim$$10$ machines.

By comparison, using the conventional \TD\ method to achieve the same latency costs
$4.9 \times 10^{13}$~\flops\ for this particular sub-bank, and so simply scaling up by the factor of $150$ suggests that it would require $7.4 \times 10^{15}$~\flops\
to search the full parameter space.  To account for the varying sample rate and template duration across the parameter space, we can also directly calculate the cost for the full \TD\ method search using expression~(\ref{eq:td-flops}), resulting in $3.8 \times 10^{15}$~\flops, agreeing within an order of magnitude.  This would require~$\gtrsim$$10^5$ current-day machines.  Presently, the \LIGO\ Data Grid%
\footnote{\url{https://www.lsc-group.phys.uwm.edu/lscdatagrid/}} consists of
only $\sim$$10^4$ machines, so direct \TD\ convolution is clearly impractical.

The overlap-save \FD\ method is slightly more efficient than LLOID for this particular sub-bank, requiring $5.2 \times 10^8$~\flops.  The scaling argument projects that a full \FD\ search would require $7.8 \times 10^{10}$~\flops.  The direct calculation from expression~(\ref{eq:fd-flops}) gives $5.9 \times 10^{10}$~\flops, in order-of-magnitude agreement.  It appears that in this application the conventional \FD\ search is scarcely a factor of 2 more computationally efficient than LLOID, but only at the price of thousands of seconds of latency.
